---
title: <center>Tutorial for merging a WRF output (maximum temperature) with ground observations using `VARMER`</center>
author: 
- Jacinto Ulloa^[jacintoisrael.ulloa@kuleuven.be]
- Esteban Samaniego^[esteban.samaniego@ucuenca.edu.ec]
- Lenin Campozano^[lenin.campozano@epn.edu.ec]
- Daniela Ballari^[dballari@uazuay.edu.ec]
- Leandro Robaina^[rm537lra@gmail.com]
- Ronald Gualán^[ronald.gualan@ucuenca.edu.ec]
date: <center>`r format(Sys.time(), '%d %B %Y')`</center>
output: 
  pdf_document:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Tutorial for merging a WRF output (maximum temperature) with ground observations using `VARMER`}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# About
This vignette describes a basic application of the `VARMER` library, which provides methods to combine satellite-based or model-based gridded images with ground-based observed time-series [1].

# Installation

Install the latest stable version (from CRAN):

```{r Install_Stable, eval = FALSE}
install.packages("VARMER")
```

# Setting up the environment

1. Load other packages that will be used in this analysis and `VARMER`:

```{r Loading_other_pks, eval = TRUE, message=FALSE}
library(zoo)
library(sf)
library(raster)
library(tictoc)
library(cluster)
library(parallel)
#library(ggplot2)
library(VARMER)
```

# Loading input data

First, daily time series of maximum temperature observations in 34 stations located in Ecuador will be used for this example, from *2004-01-01* to *2004-01-10*, which are available in the **ecuador.tmax.zoo** dataset provided in the `VARMER` package (for your own application, this dataset might be read from a CSV file or a `zoo` file). In addition, the **ecuador.tmax.stations.df** dataset contains information about the IDs and spatial coordinates of each station (for your own application, this dataset might be read from a CSV file). 
   
```{r Loading_GroundObservarions, eval = TRUE}
data(ecuador.tmax.zoo)
data(ecuador.tmax.stations.df)
```

Secondly, we need to load the WRF output. For this example, the model-data was obtained from a WRF v3.6.1 simulation at 10Km spatial resolution and at daily temporal resolution. 

   
```{r LoadingSatelliteData, eval = TRUE}
data(ecuador.tmax.wrf.out)
```

# Basic exploratory data analysis

First, we would like to visualize the first six rows of the spatial metadata:

```{r ExploringMetadata}
head(ecuador.tmax.stations.df)
```

Plotting the daily precipitation time series for the first station (code: *M0001*).

```{r PlottingPts, fig.width = 7, fig.height = 3.5, fig.align = "center", message=FALSE}
main <- paste("Daily maximum temperature for the station", ecuador.tmax.stations.df$CODIGO[1])
ylab <- "Maximum temperature [°C]"
x.ts <- ecuador.tmax.zoo[,1]

plot(x.ts, main=main, ylab= ylab, col="blue")
grid()
```

Plotting the average maximum temperature from WRF output, and overlying the boundaries of the study area (only its first attribute):

```{r PlotingTotalP, fig.width= 4, fig.height = 4, fig.align = "center", message=FALSE}
wrfout.total <- mean(ecuador.tmax.wrf.out, na.rm=FALSE)

plot(wrfout.total, main = "WRF output [Jan-2004] ", xlab = "Longitude", ylab = "Latitude")
```

# Running `VARMER`

## Setup

Finally, you need to define the output directory (`drty.out`) before running `VARMER`. Then, you can run the `varmer.ts` function as follows:

```{r, eval = FALSE}
varmer.ts(x=ecuador.tmax.zoo, x.metadata=ecuador.tmax.stations.df,
          v=ecuador.tmax.wrf.out, lat='LAT', lon='LON',
          drty.out="~/Documentos/dataset_ecuador")
```

\newpage

## Expected outputs

If `VARMER` runs without problems, the final merged product (individual *GeoTiff* files) will be stored in your user-defined `drty.out` directory.


# Software details

This tutorial was built under: 

```{r SoftwareDetails, echo=FALSE, eval = TRUE}
sessionInfo()$platform
sessionInfo()$R.version$version.string 
paste("VARMER", sessionInfo()$otherPkgs$VARMER$Version)
```


# References

1. Ulloa, J., Samaniego, E., Campozano, L., & Ballari, D. (2018). A variational merging approach to the spatial description of environmental variables. Journal of Geophysical Research: Atmospheres, 123. https://doi.org/10.1002/2017JD027982.

